Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-03 documents/ms.tex
editing the conclusion

We used the \naigrain\ simulated \kepler-like light curves for solid-body
rotation from \citet{Aigrain2015} and attempted to recover the rotation
periods used to generate them.
Three methods were compared: a Lomb-Scargle periodogram method, an
autocorrelation function method and a new Gaussian process method.
We demonstrate that the GP method produced the most precise and accurate
rotation periods of the three techniques.
As expected, it provides a large improvement over the periodogram method and a
moderate one over the ACF method.
We find that the ACF method often slightly under-predicts the rotation period
due to a subtlety in the standard peak detection algorithm.
% The periodogram method performs poorly because a sinusoid is not a perfect
% model choice for stellar light curves --- they do not always display
% sinusoidal behaviour.
% On the other hand, the ACF method does not rely on any model choice and
In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2014}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data.

The GP method provides approximate uncertainties because it produces posterior
PDF samples but they are still imperfect estimates of the intrinsic
uncertainties.
This is because the posterior PDF of the rotation period parameter is, in many
cases, multi-modal.
If it were possible to run an MCMC sampler for an infinite number of steps,
the entire posterior PDF would eventually be sampled, producing an accurate
uncertainty measurement.
We find that only a quarter of the reported uncertainties are representative.

The main aim of this work was to develop a rotation period inference method
that is probabilistic.
Probabilistic periods methods are necessary for hierarchical Bayesian
inference and for performing population analysis.
Using this method to generate a catalogue of probabilistic rotation periods
could, for example, reveal the period distribution of (and therefore
potentially the age distribution) of stars in the Milky Way.
There is still some work to be done on the development of this model such that
the uncertainty estimates are truly representative, but this new method
provides a improvement upon previous methods in that regard.

>>>>

We attempted to recover the rotation periods of \naigrain\ simulated
\kepler-like light curves for solid-body rotators in \citet{Aigrain2015} using
three different methods: a new Gaussian process method, a Lomb-Scargle
periodogram method and an autocorrelation function method.
We demonstrate that the GP method produces the most accurate rotation periods
of the three techniques, providing a large improvement over the LS periodogram
method and a moderate one over the ACF method.
We also find that the standard version of the ACF method, most commonly
implemented in the literature, often slightly under-predicts the rotation
period due to a subtle feature of its peak detection algorithm.
In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2014}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data.

Unlike the ACF and LS periodogram methods, the GP method provides posterior
PDF samples which can be used to estimate rotation periods uncertainties.
Although an improvement on competing methods, these uncertainties are still
approximate because is many cases the posterior PDFs of the parameters are
multi-modal.
In addition, although the GP model used here is clearly a good one, it is
still only an effective model, not an accurate physical model.
It can only capture the posterior PDF for the periodic component of the
covariance matrix, not the rotation period of a physical star.
If it were possible to run an MCMC sampler for an infinite number of steps,
the entire posterior PDF would eventually be sampled, and a more accurate
uncertainty would result.
Unfortunately, due to realistic computational resources and the limitations of
MCMC sampling techniques for multi-modal posteriors, we are not always able to
map out the entire posterior PDFs.
We find that only a quarter of the reported uncertainties are accurate.

The main aim of this work was to develop a rotation period inference method
that is probabilistic.
Probabilistic periods are necessary for hierarchical Bayesian inference,
particularly when performing population analysis.
This new GP method is capable of generating a catalogue of probabilistic
rotation periods which could, for example, reveal the period distribution of
(and therefore potentially the age distribution) of stars in the Milky Way.
There is still room for improvement of this model since that the uncertainty
estimates are not yet truly representative and it is still only an `effective'
model, not a physical one, but it is probabilistic, it provides more accurate
periods and, we argue, is the best method for rotation period inference
currently available.
